Estimating duration in the presence of misclassification

### 2024-03-09 introduction-comments.tex
Overleaf comments

No single metric summarises the full state of an epidemic, and no single data source fully informs all metrics of interest~\autocite{royalSocietyRnumber,pellisEstimation,paragGrowthRates}.
Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the danger of a pathogen (a pathogen is any biological agent that can cause disease).

@@

The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not capture the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is at least as important as how it is changing~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the burden of the disease, for example, the future number of severe events (hospitalisations or deaths).

\todo[inline]{What is a good definition of ``transmission'' as a concept?}
The \emph{transmission} of an infectious disease generates new infections.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether an individual is \emph{prevalent} (has the disease) is normally measurable.
Therefore, the prevalence of the population can be estimated with a prevalence survey.
However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
\cref{E-inc-prev:sec:definitions} provides a formal definition of incidence and prevalence.
Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).

To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.
% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}

However, incidence and transmission and prevalence, are almost never directly observed.
Therefore, they need to be estimated from other data sources.

Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.

@@

A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK.
Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing provides information on a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

@@

Furthermore, prevalence surveys are noisy because they are based on population sampling.
The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

@@

Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were mainly based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

>>>>

No single metric summarises the full state of an epidemic, and no single data source fully informs all metrics of interest~\autocite{royalSocietyRnumber,pellisEstimation,paragGrowthRates}.
Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the risk it poses.

@@

The most common reproduction number is the \emph{basic reproduction number}, $\R$.
$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
A fully susceptible population means none of its members have any immunity to the disease.
Note that this definition implies $\R$ is constant for a given pathogen and population.
In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R < 1$~\autocite[76]{diekmannMathematical}.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is at least as important as how it is changing~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to estimating the parameters of this process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
Therefore, the prevalence of the population can be estimated with a prevalence survey.
However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).

% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}

However, incidence and transmission and prevalence, are almost never directly observed.
Therefore, they need to be inferred from other data sources.

@@

In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.

Community testing was broadly available in England for much of the SARS-CoV-2 epidemic.
However, it does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Seroprevalence surveys estimate the proportion of the population previously infected.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) whose representativenss is unclear.

@@

Furthermore, prevalence surveys are noisy because they are based on population sampling.
The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
% I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

@@

Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
For example, the decision at each stage of lifting the lockdown in 2021 was informed by a set of modelling studies informing the government's decision~\autocite{sageEvidence}.

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
